draft

### 2025-05-23 draft.tex
Updates from Overleaf

Historically, the human eyes and brain drew conclusions from biological images but this approach is limited in scale, objectivity, and reproducibility. Now, quantification is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
We use the terms "features" and "measurements" interchangeably throughout this paper.

Image-based profiling--also termed morphological profiling--is a technique of measuring an array of shape and intensity features for all biological objects, such as cells. These features are fed into statistical or machine learning methods to identify biologically meaningful patterns. One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object"). It is commonly used to featurize the pairwise combination of all the available channels and objects. The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
\end{figure}

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, for computer-savvy scientists, its graphical user interface and lack of programmatic access means friction in pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.

From the data-science perspective, there are multiple challenges lessening the attractiveness of CellProfiler. First, coupling CellProfiler with other is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. Its usage also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} . However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}, potentially impacting the biological interpretations. Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract engineered features while readily integrating with computational workflows to achieve reproducible data analyses within a few lines of code, whether locally or in the cloud. We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem. With all this in mind we developed cp\_measure.

@@

By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility. We aim to make the image-based profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent since they are freed from all the orchestration code, we reduced the amount of time and human effort required to integrate it into existing and new pipelines, while still retaining the standard features already extracted for multiple public datasets.

@@

We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images. These correspond to 150 perturbations from the JUMP Cell Painting dataset\citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for a given feature each. We segmented images to obtain the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler, we calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.

The validation of cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}. We find that most features had a $R^2$ correlation larger than 90\%. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of features from our implementation. Some data points fall outside the diagonals for a small set of f, which hint that some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our features have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm. This result provides indicates that cp\_measure can be used in some cases instead of CellProfiler, and we would not expect a massive change in results.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocyte 3D nuclei data}
\label{sec:org447090b}

To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow. We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalinin3DCellNuclear2018}. We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells. We achieved 87\% classification accuracy in the testing dataset. Using the image-based profiles, we identified which features distinguish cells based on the time at which the image was taken. Finally, we calculated the SHAP values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.}

SHAP values of the most important features for classifying the day of differentiation (out of three days). 
Each point represents a single 3D image containing multiple nuclei. 
Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. 
The test data accuracy is shown in bold.}
\end{figure}

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. This indicates that the minor axis length of the nuclei is an indicator of phenotypic effect over the course of the experiment, implying that nuclei becomes thicker (though not necessarily more round). Though unsurprising, as we expect astrocyte nuclei to extend over time, by using cp\_measure was able to uncover this with relatively few lines of code.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these features as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological features yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

>>>>

Historically, the human eyes and brain drew conclusions from biological images but this approach is limited in scale, objectivity, and reproducibility. Now, quantification is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
We use the terms "features" and "measurements" interchangeably here.

Image-based profiling--also termed morphological profiling--is a technique of measuring an array of shape and intensity features for a population of biological objects, such as cells. These features are fed into statistical or machine learning methods to identify biologically meaningful patterns. One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object"). It can featurize the pairwise combination of all the available channels (colours) and objects. The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
\end{figure}

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, for computer-savvy scientists, its graphical user interface and lack of programmatic access means friction in pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance. For example, coupling CellProfiler with other is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. The usage of CellProfiler also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} . However, these have limitations: First, their implementation is completely independent from CellProfiler and thus they do not produce the same feature set as CellProfiler, precluding matching to existing datasets such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}. Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset. Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract engineered features while readily integrating with computational workflows to achieve reproducible data analyses with a few lines of code, whether locally or in the cloud. We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem. With all this in mind, we developed cp\_measure.

@@

By isolating the feature calculations from the graphical interface and orchestration components, we aimed to make them accessible to the data science community while improving reusability, testability and extensibility. This reduces the amount of time and human effort required to integrate featurization into existing and new pipelines, while still retaining the standard set of features available for multiple public datasets.

@@

We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images. These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature. We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction. Next, we applied cp\_measure on these masks and the original images. Finally, we mapped the features from cp\_measure to CellProfiler and calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.

We find that most features had an $R^2$ correlation larger than 90\% (Figure \ref{fig:cp_vs_cpmeasure}). We examined several individual features, typically finding straight lines demonstrating the recapitulation of features from our implementation. For a small set of features, some data points fall off the diagonal, which hint that some edge cases are treated differently by either tool; we plan to investigate these further as part of a larger project to implement unit testing in CellProfiler. These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocyte 3D nuclei data}
\label{sec:org447090b}

To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow. We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalinin3DCellNuclear2018}. We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells. We achieved 87\% classification accuracy in the testing dataset. Using the image-based profiles, we identified which features distinguish cells based on the time at which the image was taken. Finally, we calculated the SHAP values \citep{sundararajanManyShapleyValues2020} to get a better understanding of the effects of the drugs on the cells.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features refer to the nucleus and intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover with relatively few lines of code.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these features as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological features yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.
